---
title: "sc_glmpd"
author: "Jared Slosberg"
date: "2/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This workbook demonstrates the workflow and data structures associated with performing projection drivers with a GLM-based approach.

```{r init, message = F, warning=F}

#These will be loaded with the package once it is installed
#Directories are relative to this .Rmd file in vignettes folder
source("../R/model_fitting.R")
source("../R/visualization.R")


require(monocle3)
require(purrr)
require(tidyr)
require(tibble)
require(dplyr)
require(MASS)
require(ggplot2)
require(furrr)
require(ComplexHeatmap)
require(RhpcBLASctl)




```

```{r import}
#dataset to project into patterns
cds <- readRDS("../inst/extdata/6month_LMMP.rds")

#p-weights resulting from projection with projectR
#rownames are cell barcodes, column names are pattern names. eg "cellPattern1"
projected_weights <- read.csv("../inst/extdata/6mo_LMMP_cell_size_log_in_TC_LMMP_9-18_patterns.csv", row.names = 1)


```

```{r prepare_cds}
#pattern # to project
select_patterns <- c(27,43)
projected_weights <- projected_weights[paste0("cellPattern", select_patterns)]

#remove lowly expressed genes that will not result in good fits
fData(cds)$num_transcripts <- Matrix::rowSums(exprs(cds))
min_n_transcripts <- 150
min_pct_cells_expressing <- .005

min_cells_expressing <- min_pct_cells_expressing*dim(cds)[2]
genes_to_use <- which(fData(cds)$num_transcripts > min_n_transcripts & fData(cds)$num_cells_expressed > min_cells_expressing)
#character vector of gene names
genes_of_interest <- fData(cds)[genes_to_use, "gene_id"]

cds_sub <- cds[fData(cds)$gene_id %in% genes_of_interest,]
#TODO: Make sure gene short names are all unique

#downsample for ease of running the vignette
set.seed(208)
downsample <- 1
if(downsample){
  num_genes <- 200
  down_gene <- sample(dim(cds_sub)[1], num_genes)
  num_cells <- 2000
  down_cell <- sample(dim(cds_sub)[2], num_cells)
  
  cds_sub <- cds_sub[down_gene,down_cell]
  
  #character vector of final gene names
  genes_of_interest <- fData(cds_sub)[,"gene_id"]
}

cds_sub
```

```{r run_model_fit, warning = F}
#patternWeight is hard-coded in the fit_models_cds function to be the p-weights column provided
model_str <- paste0("~0 + patternWeight:cell_type + cell_type")

#specify parameters for fittings
exp_family <- "negbinomial"
#capable of parallelization over patterns (columns of provided projection matrix)
ncores = 1
#return the full glm output
return_option <- "full_model"

system.time(glm_models_list <- fit_model_cds(cds = cds_sub, model_formula_str = model_str, projected_patterns = projected_weights,
                                          exp_family = exp_family, cores = ncores, clean_model = F, verbose = T, result = return_option))

class(glm_models_list)
length(glm_models_list)
names(glm_models_list)

#model results for first pattern
purrr::pluck(glm_models_list, paste0("cellPattern",select_patterns[1]))
```

Model fittings are complete, now we can extract the relevant info from the models using extractCoefficients(). The first order list is always the patterns. Provided the genes used in the fittings above, it generates a data frame for each gene with each row representing one of the model parameters, and each column representing a measurement (eg. estimate, p-value...). 
```{r extract_model_coefficients}
#genes vector is used to pull parameter estimates from the model info for each gene. Must be the same as provided in the cds_sub that is fitted
#organizes each pattern such that one list is present for each gene, with all coefficients together
extracted_coefficients_list <- extractCoefficients(glm_models = glm_models_list, genes = genes_of_interest)

names(extracted_coefficients_list)
head(names(extracted_coefficients_list[[1]]))

#for one pattern, for one gene
purrr::pluck(extracted_coefficients_list, paste0("cellPattern",select_patterns[1]), genes_of_interest[1])

```

```{r order_by_gene}

gene_coefficients_list <- orderCoefficientsByGene(extracted_coefficients_list, filter_significance = NULL, string = NULL)
gene_coefficients_list

```






